Graph Topological Aspects of Granger Causal Network Learning

### 2019-07-23 paper_theory/paper_theory.tex
Substantial updates to the proof of the confounding theorem.

% \theoremstyle{definition}
\newtheorem{remark}{Remark}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{assumptions}{Assumptions}

@@

The aforementioned assumptions are quite weak.  The strongest
assumption we require is finally that $\Sigma_v$ is a diagonal matrix,
which is referred to as a lack of instantaneous feedback in $x(t)$.
We formally state our setup as a definition, which is the setup for
the remainder of the paper:

\begin{definition}[Basic Setup]
  \label{def:basic_setup}
  The process $x(t)$ is an $n$ dimensional wide sense stationary
  process having invertible $\VAR(\infty)$ representation
  \eqref{eqn:ar_representation} where $v(t)$ is sequentially
  uncorrelated and has a diagonal covariance matrix.  The $MA(\infty)$
  representation of equation \eqref{eqn:wold} has $c(t) = 0$ and
  $A(0) = I$.
\end{definition}

@@

\subsection{Persistent Systems}
\label{sec:persistent_systems}
In section \ref{sec:strongly_causal_graphs} we obtained a converse to
part $(a)$ of Proposition \ref{prop:ancestor_properties} via the
notion of a strongly causal graph topology.  In this section, we
complete a converse by adding the additional requirement we refer to
as ``persistence''.

@@

The interpretation of the following persistence condition is that
each node stores some ``memory'' of the past.

@@

\begin{remark}
  In the context of Granger-causality, ``most'' systems should be
  persistent.  In particular, $\mathsf{\VAR}(p)$ models are likely to
  be persistent since these naturally result in an equivalent
  $\mathsf{MA}(\infty)$ representation, except for the pathological case
  where $\B(z)$ is Nilpotent.  Indeed, in a $\VAR(1)$ model, if the
  system matrix is diagonalizable and has at least two distinct
  eigenvalues, then the system is persistent.  See Example
  \ref{ex:persistent_system}.

  Moreover, persistence is not the weakest condition necessary for the
  results of this section, the condition
  $\tau_0(\A_{jk}) < \tau_\infty(\A_{ik})$ for each $i, j, k$ such that
  $k \in \anc{i} \cap \anc{j}$ is enough.  The intuition being that nodes
  $i$ and $j$ are not receiving temporally disjoint information from
  $k$.
\end{remark}

\begin{example}
  \label{ex:persistent_system}
  Consider a process $x(t)$ generated by the $\VAR(1)$
  model\footnote{Recall that any $\VAR(p)$ model with $p < \infty$ can be
    written as a $\VAR(1)$ model, so we lose little generality in
    considering this case.}  having $\B(z) = Bz^{-1}$, we will examine
  conditions that guarantee persistence.  Pick any
  $i \in [n], j \in \anc{i} \setminus \{i\}$.  Then the stability of
  $B$ allows us to write

  \begin{equation*}
    \A(z) = \sum_{k = 0}^\infty B^k z^{-k},
  \end{equation*}

  whereby we see that $\exists k > 0$ such that $[B^k]_{ij} \ne 0$
  (since $j \in \anc{i}$).  Then consider

  \begin{equation*}
    \begin{aligned}
      [B^{rk}]_{ij} &= e_i^\T B^{rk} e_j \\
      &\overset{(a)}{=} \big((P^\T e_i)^\T J^{rk} P^{-1}e_j\big)\\
      &= \tr [(P^\T e_i)^\T J^{rk} P^{-1}e_j]\\
      &\overset{(b)}{=} \tr [(J^{rk}) (v u^\T)],
    \end{aligned}
  \end{equation*}

  where $(a)$ utilizes the Jordan Normal Form of $B$, and $(b)$
  denotes $u = P^\T e_i$ and $v = P^{-1}e_j$.  In order for
  $\tau_\infty(\A_{ij}) < \infty$, there must be some $N > 1$ such
  that $\forall r \ge N$, the above term is $0$.  This may be the case
  for instance if $B$ is a nilpotent matrix.

  Let us suppose now that $B$ is diagonalizable (i.e. $J$ is a
  diagonal matrix) with at least $2$ distinct eigenvalues; in this
  case $B$ is also \textit{not} nilpotent.  We can then rewrite the
  above as

  \begin{equation*}
    f(r) \defeq \tr [(J^{rk}) (v u^\T)] = \sum_{\nu = 1}^n \lambda_\nu ^{rk} v_\nu u_\nu \defeq \sum_{\nu = 1}^n \lambda_\nu^{rk} \beta_\nu
  \end{equation*}

  where $\lambda_\nu$ denotes the eigenvalues of $B$ and
  $\beta_\nu = u_\nu v_\nu$.  Note that $f(0) = 0$ since $i \ne j$ and
  $u$ is a row of $P$ and $v$ is a column of $P^{-1}$.  Moreover,
  $f(1) \ne 0$ by hypothesis.  But, in order for
  $f(r) = 0\ \forall r \ge N$, it would need to be the case that

  \begin{equation*}
    \Dg(\bm{\lambda})^r \bm{\lambda} = Vz
  \end{equation*}

  had a solution in $z$ for every $r \ge N$, where $V$ is an
  $n \times n - 1$ full-rank matrix whose columns span the nullspace of $\beta$,
  and $\bm{\lambda} = (\lambda_1, \ldots, \lambda_n)$. That is,
  iterates of $\Dg(\bm{\lambda})$ applied to $\bm{\lambda}$ would need to remain
  inside $\beta$'s nullspace.  This would imply that

  \begin{equation*}
    VV^\dagger \bm{\lambda}^{r + 1} = \bm{\lambda}^{r + 1},
  \end{equation*}

  i.e. that $\bm{\lambda}^{r + 1}$ is an eigenvector of $VV^\dagger$
  for an infinite number of integers $r$ (the exponentiation is to be
  understood as a point wise operation).  However, since there can only
  be a finite number of (unit length) eigenvectors, this cannot be the
  case unless every eigenvalue $(\lambda_1, \ldots, \lambda_n)$ were
  equal.

  We see from this example that the collection of $\VAR(1)$ systems
  which are not persistent are pathological, in the sense that their
  system matrices have zero measure when viewed as a subset of $\R^{n^2}$.
\end{example}

\begin{proposition}
  \label{prop:persistence_converse}
  Suppose $\gcg$ is a strongly causal DAG and that $x(t)$ is
  persistent, then if there exists a $k$ which confounds $(i, j)$ we
  have $i \pwgc j$ and $j \pwgc i$.
\end{proposition}

\begin{remark}
  The importance of this result is that it allows us to eliminate the
  effects of confounding by removing bidirectional edges.
\end{remark}

@@

\begin{theorem}[Pairwise Recovery]
  \label{thm:scg_recovery}
  If the Granger-causality graph $\gcg$ for persistent process $x(t)$
  is a strongly causal DAG then $\gcg$ can be inferred from pairwise
  causality tests.  The procedure can be carried out, assuming
  we have an oracle for pairwise causality, via Algorithm
  (\ref{alg:pwgr}).
\end{theorem}

@@

The theorem is proven in the supplementary material by establishing
the correctness of Algorithm (\ref{alg:pwgr}).  The idea is to
iteratively ``peel away layers'' of nodes by removing the nodes that
have no parents remaining.  The requirement of strong causality
ensures that all actual edges of $\gcg$ manifest in some way as
pairwise relations (by Proposition \ref{prop:pwgc_anc}), and the
requirement of persistence allows confounding to be eliminated by
removing bidirectional edges(by Proposition
\ref{prop:persistence_converse} and Corollary
\ref{cor:bidirectional_edge}).  Without persistence, each confounded
pair would give rise to $4$ possible pairwise topologies consistent
with $\gcg$, one for each type of pairwise edge (no edge,
unidirectional, bidirectional).

>>>>

% \theoremstyle{definition}
\newtheorem{remark}{Remark}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{assumption}{Assumption}

@@

The aforementioned assumptions are quite weak.  The strongest
assumption we require is finally that $\Sigma_v$ is a diagonal
positive-definite matrix, which is referred to as a lack of
instantaneous feedback in $x(t)$.  We formally state our setup as a
definition, which is the setup for the remainder of the paper:

\begin{definition}[Basic Setup]
  \label{def:basic_setup}
  The process $x(t)$ is an $n$ dimensional wide sense stationary
  process having invertible $\VAR(\infty)$ representation
  \eqref{eqn:ar_representation} where $v(t)$ is sequentially
  uncorrelated and has a diagonal positive-definite covariance matrix.
  The $MA(\infty)$ representation of equation \eqref{eqn:wold} has
  $c(t) = 0$ and $A(0) = I$.
\end{definition}

@@

\subsection{Persistent Systems}
\label{sec:persistent_systems}
In section \ref{sec:strongly_causal_graphs} we obtained a converse to
part $(a)$ of Proposition \ref{prop:ancestor_properties} via the
notion of a strongly causal graph topology (see Proposition
\ref{prop:pwgc_anc}).  In this section, we study conditions under
which a converse to part $(b)$ will hold.

@@

@@

\begin{remark}
  In the context of Granger-causality, ``most'' systems should be
  persistent.  In particular, $\mathsf{\VAR}(p)$ models are likely to
  be persistent since these naturally result in an equivalent
  $\mathsf{MA}(\infty)$ representation, see Example
  \ref{ex:persistent_system}.

  Moreover, persistence is not the weakest condition necessary for the
  results of this section, the condition that for each $i, j$ there is
  some $k \in \anc{i}\cap\anc{j}$ such that
  $\tau_0(\A_{jk}) < \tau_\infty(\A_{ik})$ is enough.  The intuition
  being that nodes $i$ and $j$ are not receiving temporally disjoint
  information from $k$.

  The etymology for the persistence condition can be explained by
  supposing that the two nodes $i, j$ each have a loop (i.e.
  $\B_{ii}(z) \ne 0, \B_{jj}(z) \ne 0$) then this autoregressive
  component acts as ``memory'', and so the influence from the
  confounder $k$ \textit{persists} in $x_i(t)$, and
  $\tau_\infty(\A_{ik}) = \infty$ for each confounder is expected.
\end{remark}

\begin{example}
  \label{ex:persistent_system}
  Consider a process $x(t)$ generated by the $\VAR(1)$
  model\footnote{Recall that any $\VAR(p)$ model with $p < \infty$ can
    be written as a $\VAR(1)$ model, so we lose little generality in
    considering this case.}  having $\B(z) = Bz^{-1}$.  If $B$ is
  diagonalizable, and has at least $2$ distinct eigenvalues, then
  $x(t)$ is persistent.

  See the supplementary material for an analysis of this example.
\end{example}

In order to eliminate the possibility of a particular sort of
cancellation, an ad-hoc assumption is required.  Strictly speaking,
the persistence condition is not a necessary or sufficient condition
for the following, but cases where the following fails to hold, and
peristence \textit{does} hold, are unavoidable pathologies.

\begin{assumption}[no-cancellation]
  \label{ass:T_causality}
  Fix $i, j \in [n]$ and let $H_i(z)$ be the strictly-causal filter
  such that

  \begin{equation*}
    H_i(z)x_i(t) = \linE{x_i(t)}{\H_{t - 1}^i},
  \end{equation*}

  and similarly for $H_j(z)$.  Then define

  \begin{equation}
    \label{eqn:T_filter}
    T_{ij}(z) = \sum_{k \in \anc{i} \cap \anc{j}}\sigma_k^2\A_{ik}(z^{-1})(1 - H_i(z^{-1}))(1 - H_j(z))\A_{jk}(z),
  \end{equation}

  where $\sigma_k^2 = \E v_k(t)^2$.

  We will say that Assumption \ref{ass:T_causality} is satisfied if
  for every $i, j \in [n]$, $T_{ij}(z)$ is either constant over $z$
  (i.e. each $z^k$ coefficient for $k \in \Z \setminus \{0\}$ is $0$),
  or is \textit{neither} causal (i.e. containing only $z^{-k}$ terms,
  for $k \ge 0$) \textit{or} anti-causal (i.e. containing only $z^k$
  terms, for $k \ge 0$).  Put succinctly, $T_{ij}(z)$ must be
  two-sided.
\end{assumption}

\begin{remark}
  Under the condition of persistence, the only way for Assumption
  \ref{ass:T_causality} to fail is through cancellation in the terms
  defining $T_{ij}(z)$.  For example, the condition is assured if
  $x(t)$ is persistent, and there is only a single confounder.
  Unfortunately, some pathological behaviour resulting from
  confounding nodes seems to be unavoidable without some assumptions
  about the parameters of the $MA(\infty)$ system defining $x(t)$.
\end{remark}

\begin{proposition}
  \label{prop:persistence_converse}
  Fix $i, j \in [n]$ and suppose $\exists k \in \anc{i} \cap \anc{j}$
  which confounds $i, j$.  Then, if $T_{ij}(z)$ is not causal we have
  $j \pwgc i$, and if $T_{ij}(z)$ is not anti-causal we have
  $i \pwgc j$.  Moreover, if Assumption \ref{ass:T_causality} is
  satisfied, then $j \pwgc i \iff i \pwgc j$.
  % Suppose that Assumption \ref{ass:T_causality} is satisfied.  Then if
  % there exists a $k$ which confounds $(i, j)$ we have
  % $i \pwgc j \implies j \pwgc i$.  Moreover, if $T_{ij}(z)$ is not
  % constant, then $i \pwgc j$.
\end{proposition}

\begin{remark}
  The importance of this result is that when $i \pwgc j$ is a result
  of a condounder $k$, then $i \pwgc j \iff j \pwgc i$.  This
  implies that in a strongly causal graph every bidirectional pairwise
  causality relation must be the result of a confounder.  Therefore,
  in a strongly causal graph, pairwise causality analysis is
  \textit{immune to confounding} (since we can safely remove all
  bidrectional edges).
\end{remark}

@@

\begin{theorem}[Pairwise Recovery]
  \label{thm:scg_recovery}
  If the Granger-causality graph $\gcg$ for the process $x(t)$ is a
  strongly causal DAG and Assumption \ref{ass:T_causality} holds, then
  $\gcg$ can be inferred from pairwise causality tests.  The procedure
  can be carried out, assuming we have an oracle for pairwise
  causality, via Algorithm (\ref{alg:pwgr}).
\end{theorem}

@@

The theorem is proven in the supplementary material by establishing
the correctness of Algorithm (\ref{alg:pwgr}).  The idea is to
iteratively ``peel away layers'' of nodes by removing the nodes that
have no parents remaining.  The requirement of strong causality
ensures that all actual edges of $\gcg$ manifest in some way as
pairwise relations (by Proposition \ref{prop:pwgc_anc}), and the
no-cancellation condition of Assumption \ref{ass:T_causality} allows
confounding to be eliminated by removing bidirectional edges (by
Proposition \ref{prop:persistence_converse} and Corollary
\ref{cor:bidirectional_edge}).  Without Assumption
\ref{ass:T_causality}, then each confounded pair would give rise to
$4$ possible pairwise topologies consistent with $\gcg$, one for each
type of pairwise edge (no edge, unidirectional, bidirectional).
